Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-05 paper/paper.tex
beauty

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.2$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This data-driven model contains no knowledge of stellar astrophysics or the
Milky Way, but rather derives its precision from its generative model of noisy
parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The \xd-optimized prior is used to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a posterior parallax and distance sampling for 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

>>>>

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.2$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise (and beautiful) because it is a generative model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The \xd-optimized prior is used to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a posterior parallax and distance sampling for 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}
